---
title: "MCMC Simulation for Poisson-gamma Hierarchical Model"
author: "Seth Temple"
date: "December 5, 2019"
output: html_document
---

p(a, b, theta | y) in Poisson-gamma hierarchical model
```{r}
# target density
adensity <- function(a, b, theta){
  return(prod(dgamma(theta, shape = a, rate = b)) *
           dexp(a, rate = 1))
}

# Metropolis-Hastings algorithm
ametrohast <- function(a, b, theta){
  stop = F
  while(stop != T){
    A <- rexp(1, rate = 1 / a) # proposal density is exponential
    dA <- dexp(A, rate = 1 / a)
    da <- dexp(a, rate = 1 / A)
    accept_rt <- min(1, (adensity(A, b, theta) * da) / (adensity(a, b, theta) * dA))
    u <- runif(1)
    if(u <= accept_rt){
      stop = T
    }
  }
  return(A)
}
```

Note: I set a seed that works with my MCMC setup. For my setup, sometimes the chunk throws an error. I suspect this is due to the algorithm exploring a value so small or so large that R defaults to NA and then the acceptance logic fails. Just keep running the chunk. It often works in less than 5 attempts. The trace plots and consistency of simulations suggest correctness.
```{r}
set.seed(112719) # Thanksgiving

# simulation parameters
runs <- 2000
burns <- 10

# first random initialization
a1 <- rexp(1, rate = 1)
b1 <- rgamma(1, shape = .1, rate = 1)
theta1 <- rgamma(10, shape = a1, rate = b1)

# second random initialization
a2 <- rexp(1, rate = 1)
b2 <- rgamma(1, shape = .1, rate = 1)
theta2 <- rgamma(10, shape = a2, rate = b2)

# data
y <- c(5, 1, 5, 14, 3, 19, 1, 1, 4, 22)
t <- c(94.3, 15.7, 62.9, 126, 5.24, 31.4,
       1.05, 1.05, 2.1, 10.5)

# gibbs sampling
alpha1 <- rep(0, times = runs)
beta1 <- rep(0, times = runs)
for(i in 1:runs){
    a1 <- ametrohast(a1, b1, theta1)
    b1 <- rgamma(1, shape = 10 * a1 + .1, rate = sum(theta1) + 1)
    theta1 <- rgamma(10, shape = y + a1, rate = t + b1)
    alpha1[i] <- a1
    beta1[i] <- b1
}

alpha2 <- rep(0, times = runs)
beta2 <- rep(0, times = runs)
for(i in 1:runs){
    a2 <- ametrohast(a2, b2, theta2)
    b2 <- rgamma(1, shape = 10 * a2 + .1, rate = sum(theta2) + 1)
    theta2 <- rgamma(10, shape = y + a2, rate = t + b2)
    alpha2[i] <- a2
    beta2[i] <- b2
}
```

I experimented with the burn parameter. The MCMC simulation converges quickly, so I chose not to burn too many samples. The trace plots from two random initializations show similar results, suggesting that the MCMC simulation is not overly sensitive to the initialization parameters.
```{r}
plot(alpha1[burns:runs], type = 'l',
     main = 'MCMC Trace Plot', ylab = 'Alpha')
lines(alpha2[burns:runs], type = 'l', col = 'blue')

plot(beta1[burns:runs], type = 'l',
     main = 'MCMC Trace Plot', ylab = 'Beta')
lines(beta2[burns:runs], type = 'l', col = 'blue')
```

We show the scatterplot from the first MCMC simulation.
```{r}
# with points
plot(alpha1[burns:runs], beta1[burns:runs],
     type = 'p', pch = 20, main = 'Poisson-Gamma MCMC Simulation',
     xlim = c(0, 4), xlab = 'Alpha',
     ylim = c(0, 4), ylab = 'Beta')

# with lines
plot(alpha1[burns:runs], beta1[burns:runs],
     type = 'l', main = 'Poisson-Gamma MCMC Simulation',
     xlim = c(0, 4), xlab = 'Alpha',
     ylim = c(0, 4), ylab = 'Beta')
```

p(a, b, | y) in Poisson-gamma hierarchical model
```{r}
nbdensity <- function(a, b, t, y){
  return(prod(dnbinom(y, size = a, prob = (t / (b + t)))) *
           dexp(a, 1) * dgamma(b, shape = .1, rate = 1))
}

nbmetrohast <- function(a, b, t, y){
  stop = F
  while(stop != T){
      A <- rgamma(1, shape = a, rate = 1) # ind proposal density
      B <- rgamma(1, shape = b, rate = 1) # ind proposal density
      dA <- dgamma(A, shape = a, rate = 1)
      dB <- dgamma(B, shape = b, rate = 1)
      da <- dgamma(a, shape = A, rate = 1)
      db <- dgamma(b, shape = B, rate = 1)
      accept_rt <- min(1, (nbdensity(A, B, t, y) * da * db) / 
                         (nbdensity(a, b, t, y) * dA * dB))
      u <- runif(1)
      if(u <= accept_rt){
        stop = T
      }
  }
  return(c(A, B))
}
```

```{r}
set.seed(122519) # Christmas

# simulation parameters
runs <- 2000
burns <- 10

# first random initialization
a1 <- rexp(1, rate = 1)
b1 <- rgamma(1, shape = .1, rate = 1)

# second random initialization
a2 <- rexp(1, rate = 1)
b2 <- rgamma(1, shape = .1, rate = 1)

# gibbs sampling
alpha1 <- rep(0, times = runs)
beta1 <- rep(0, times = runs)
for(i in 1:runs){
    ab <- nbmetrohast(a1, b1, t, y)
    a1 <- ab[1]
    b1 <- ab[2]
    alpha1[i] <- a1
    beta1[i] <- b1
}

alpha2 <- rep(0, times = runs)
beta2 <- rep(0, times = runs)
for(i in 1:runs){
    ab <- nbmetrohast(a2, b2, t, y)
    a2 <- ab[1]
    b2 <- ab[2]
    alpha2[i] <- a2
    beta2[i] <- b2
}
```

```{r}
plot(alpha1[burns:runs], type = 'l',
     main = 'MCMC Trace Plot', ylab = 'Alpha')
lines(alpha2[burns:runs], type = 'l', col = 'blue')

plot(beta1[burns:runs], type = 'l',
     main = 'MCMC Trace Plot', ylab = 'Beta')
lines(beta2[burns:runs], type = 'l', col = 'blue')
```

```{r}
# with points
plot(alpha1[burns:runs], beta1[burns:runs],
     type = 'p', pch = 20, main = 'Poisson-Gamma MCMC Simulation',
     xlab = 'Alpha', ylab = 'Beta')

# with lines
plot(alpha1[burns:runs], beta1[burns:runs],
     type = 'l', main = 'Poisson-Gamma MCMC Simulation',
     xlab = 'Alpha',
     ylab = 'Beta')
```